Scan-Centric, Frequency-Based Method for Characterizing Peaks from Direct Injection Fourier transform Mass Spectrometry Experiments

### 2022-05-18 doc/peakcharacterization_manuscript.Rmd
some more changes
close #50
close #42
close #41
close #40
close #37
close #36
close #34

::: {custom-style="MDPI_1.7_abstract"}
**Abstract:** We introduce a novel, scan-centric method for characterizing peaks from direct injection multi-scan Fourier transform mass spectra of complex samples that utilizes frequency values derived directly from the spacing of raw m/z points in spectral scans.
Our peak characterization method utilizes intensity independent noise removal and normalization of scan-level data to provide a much better fit of relative intensity to natural abundance probabilities for low abundance isotopologues that are not present in all of the acquired scans.
Moreover, our method calculates both peak- and scan-specific statistics incorporated within a series of quality control steps that are designed to robustly derive peak centers, intensities and intensity ratios with their scan-level variances.
These cross-scan characterized peaks are suitable for use in our previously published peak assignment methodology, Small Molecule Isotope Resolved Formula Enumeration (SMIRFE).
:::

@@

The input peak data in these studies are derived from multi-scan direct-injection FT-MS data.
SMIRFE uses expected relationships of relative peak height or intensity to natural abundance probability (NAP) across multiple peak pairs to determine the likelihood of an assignment for the peaks in question.
In the Thermo-Fisher Orbitrap FT-MS instruments, data is acquired in microscans and scans, where each scan is an aggregate of multiple microscans.
While these micro-scan and subsequent scans are expected to be analytical replicates, spray instability and temporal delay in automatic gain control pragmatically break this expectation.
Therefore, it is advantageous to keep the data at the scan level, so that any given poor quality (i.e. "bad") scan can be removed prior to aggregating the data across scans and generating a centroided peak m/z and intensity.
If the scan level data is not processed and aggregated correctly, the final peak intensities in arbitrary units and centers in m/z will have high uncertainty (total variance) [@moseleyErrorAnalysisPropagation2013].

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("motivation")`.**
**A** Xcalibur intensities (black) and theoretical intensities based on relative NAP (red) for ECF derived threonine.
The 18O isotopologue is shifted by 0.01 m/z and the NAP peaks by 0.05 m/z to aid visualization.
**B** Theoretical NAP-based-Log10-intensities and Log10 XCalibur intensities.
The red line indicates perfect agreement.
:::

@@

The constant point difference of frequency space is useful, because some of the subsequent steps in our workflow use sliding and tiled windows where it is assumed that the sliding windows contain the same number of data points.
The m/z point-to-point differences are not constant, but can be approximated by a LOESS linear model [@Cleveland1992loess]; however, it is very difficult to create a LOESS model with an intercept of 0.
In addition, we would also need to vary the width of sliding windows according to the m/z difference at a particular m/z based on the LOESS model.
Frequency-based points suffer none of these drawbacks, and the conversion from m/z can be derived from the raw profile level data itself, which is incredibly useful.
:::

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("mz_frequency_conversion")`.** 
**A**: Intensity vs m/z for a single peak from a single scan.
Red lines denote the difference between each adjacent pair of points, and red dots the m/z midpoint between the pair of points.
The difference divided by the midpoint is used to derive the frequency values in **B**.
**B** plots the intensity vs the converted frequency points derived from **A**.
The red lines denote the differenc between adjacent points, which are shown in the y-axis of **C** for this single peak.
A histogram of adjacent point differences for all points is shown in **D**.
The differences for all points in a single scan vs m/z are shown in **E**, with those differences that lie within 0.49 - 0.51 shown in red.
**F** shows the plot of derived spectral frequency vs m/z, with fitted values from the linear regression model in red.
:::

@@

```{r increment_slidingwindow}
figure_count$increment("slidingwindow_count")
tar_load(mn_ratios)
tar_load(sliding_regions)
tar_load(noise_plot)
```

::: {custom-style="MDPI_3.1_text"}
In a dataset of this nature, we expect that much of the data is really just noise, and doesn't contribute that much to the analysis.
Furthermore, it is expected that noise is randomly distributed across the scans.
Therefore, if we slide a window across the data and sum the number of non-zero points in each window, we expect that most of the data we encounter is actually noise.
Subsequently, we divide the counts into tiled regions of a set size (2000 in this work), and examine the 99th percentile of the sliding window counts within the tiled region.
A histogram of sliding window counts in a single tiled region is shown in `r figure_count$label_text("slidingwindow_count")`A.
The distribution of 99th percentiles across all of the tiled regions for a single sample is shown in `r figure_count$label_text("slidingwindow_count")`B.
We can use the median (50th percentile) value of all the 99th percentiles from all of the tiled regions to determine a cutoff value above which any non-zero points are potentially **signal** and not just noise.
In the example below, the median of the tiled windows is 5, resulting in a cutoff of 7.5.
The number of peaks detected without applying a cutoff explodes anywhere from `r mn_ratios[["97lipid"]]$ratio`-fold (from `r format(mn_ratios[["97lipid"]]$count_99, big.mark = ",")` to `r format(mn_ratios[["97lipid"]]$count_0, big.mark = ",")` in the 97lipid sample), to `r mn_ratios[["2ecf"]]$ratio`-fold (from `r format(mn_ratios[["2ecf"]]$count_99, big.mark = ",")` to `r format(mn_ratios[["2ecf"]]$count_0, big.mark = ",")` in the 2ecf sample) with a vast majority of the peaks being present in a single scan.
By discarding those windows with a non-zero point density below 1.5x the median of 99th percentiles, and merging the remaining adjacent windows, the regions most likely to contain actual, informative peaks are rapidly and easily determined.
:::

@@

```{r slidingwindow_count, dn_id = figure_count, fig.width = 8, fig.height = 8}
(sliding_regions / noise_plot) + plot_annotation(tag_levels = "A")
```

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("slidingwindow_count")`.**
**A**: Histogram of the number of non-zero points from sliding windows ten frequency points wide and a stride of one point, from a single tiled window 2000 points wide.
Blue line is the 99th percentile.
**B**: Histogram of the 99th percentile cutoffs from all of the tiled windows.
Red line is the median x 1.5.
**C** and **D** show the number of initial regions as a function of the percentile cutoff used for either the AA ECF (C) or lipid (D) samples.
:::

@@

::: {custom-style="MDPI_3.1_text"}
Although many other types of mass spectrometry data suffer from a variable and noisy baseline, the scan-level profile data from the Thermo-Fisher Fusion has a baseline of 0 due to manipulations in the Thermo-Fisher firmware, making the determination of the centroided values considerably easier. 
For each region initially created, the peaks in each scan within that region can be characterized (i.e. centroiding).
For centroiding, we use a simple weighted quadratic model of log(intensity) to m/z. 
We add a small constant to enable using the zero values, and weight the values by their ratio to the most intense value, which is normally the value closest to the center of peak, helping to ensure that the **true** centroid is determined. 
From the fitted model, we can derive the centroided center and the intensity of the peak, as shown in `r figure_count$label_text("centroided_peaks")`.
:::

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("breaking_regions")`.** Splitting a single region into two regions based on the peaks that are present.
**A**: The full set of raw frequency and intensity data across all scans for the region are shown.
Clearly the region has two separate peaks within it.
**B**: The peak centroids (frequency and intensity) for each peak in black.
The tiled regions (red) are used to quantify the number of peaks.
**C**: The number of peaks within each tiled region shown as a histogram.
Each group of non-zero adjacent regions will be merged to form a new peak region.
:::

@@

::: {custom-style="MDPI_3.1_text"}
Due to differences in how many ions are captured in the trap and the limited dynamic range of the detector, the observed peak intensities for the same analyte vary between scans (see `r figure_count$label_text("breaking_regions")`A and B for example).
Using the median peak differences between scans, it is possible to normalize the peak intensities across scans.
However, there are two issues with these peak intensities across scans: (a) some peak intensities correlate with the scan number (i.e. scan acquisition order); and (b) some peak differences between scans are correlated with intensity.
The solution to (a) is to do a two-pass normalization.
After the first pass, the peaks whose intensity is correlated with scan order are detected (absolute Pearson correlation > 0.5).
In the second pass, the correlated peaks are removed, and normalization is carried out again.
`r figure_count$label_text("intensity_scan")`A shows an example peak whose intensity across scans is correlated with scan number.
The solution to (b) (peak differences correlated with intensity) is to only use the most intense peaks, as shown in `r figure_count$label_text("intensity_scan")`B.
The highlighted peaks in `r figure_count$label_text("intensity_scan")`B are those with an intensity greater than 0.7 of the maximum intensity observed in that scan, and at least visually, their differences are **not** correlated with intensity.
If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in `r figure_count$label_text("normalization_factors")`A and B.
Though the two-pass normalization (doublenorm) and single-pass normalization using the most intense peaks (singlenorm_int) generate very similar normalization factors (`r figure_count$label_text("normalization_factors")`B and C), removing the scan-correlated peaks does change some of the normalization factors.
:::

@@

::: {custom-style="MDPI_3.1_text"}
We have previously described the presence of high peak density (HPD) artifacts in FT-MS spectra [@mitchellNewMethodsIdentify2018].
Ideally, the peak characterization procedure should reduce and mitigate their presence and effect in the resultant reported peaks.
Their presence should be minimized by removing noise peaks, and removing peak regions that have multiple reported peaks from the same scan.
However, we expect they may also present as characterized peaks that have larger than expected frequency level standard deviations (FSD) when calculated across scans.
These peaks can be detected by simply examining the distribution of FSDs and marking outlier peaks.
To verify the mitigation of HPD regions, we converted centroided m/z's from XCalibur to frequency values using the previously calculated model for that sample, and measured peak density to detect HPD regions, and compare them with the scan-centric peaks and FSD outliers.
XCalibur peak density was measured using a sliding window ten points wide and a stride of one point.

@@

::: {custom-style="MDPI_3.1_text"}
Each step in the peak characterization either changes the overall number of peaks coming from each scan (sliding windows and breaking initial regions) or the overall intensity of the points within a scan.
Therefore, one way to quantify any potential *improvements* in the characterized peaks is to look at the relative standard deviation (RSD) for the characterized scan level peak intensities (calculated as the standard deviation of peak heights across scans divided by the mean peak height) and compare them as each processing step is introduced.
`r figure_count$label_text("rsd_method")` illustrates the peak height RSD distributions for four different samples.
Up to two modes are reported for each distribution, where a mode location is only reported if it is ≥ 0.2X the most intense mode.
For samples 1ecf and 2ecf, there is a general shift in the RSD to the left, going from the bottom processing methods (msnbase_only) to the top processing method (filtersd), representing a visually clear improvement.
For samples 49lipid and 97lipid, the improvements are visually more subtle, having a slight shift to the left as well as a narrowing and smoothing of the RSD distribution.
Part of this sublety is likely due to the bi-modality of these two distributions.
This is not surprising, since these two samples are non-polar extractions from tissue and are biochemically more complex.
`r table_count$label_text("rsd_method")` provides more quantitative metrics.
At two decimal places, the filtersd, doublenorm, and singlenorm processing methods give superior and nearly identical results for three of the four samples, especially in terms of mean and median RSD.
However, for sample 1ecf, the max RSD is much higher for singlenorm, highlighting its instability.
:::

@@

::: {custom-style="MDPI_3.1_text"}
The 97lipid filtersd RSD distribution (as well as others) is bi-modal.
`r figure_count$label_text("rsd_intensity")` plots each peaks RSD as a function of the mean intensity across scans, for every peak present in at least three scans (top), or the peaks present in at least 80% of the scans (bottom).
At least part of the bi-modality in the RSD distribution appears to be related to this dependence of RSD on intensity, as well as truncation as peaks appear in fewer scans.
This is supported by part of the distribution disappearing when we require that peaks are in at least 80% of the available scans (difference between `r figure_count$label_text("rsd_intensity")` top and bottom).

@@

::: {custom-style="MDPI_3.1_text"}
Although we expected that the corrected scan-centric intensities would behave similarly or better compared to the raw intensities, `r supp_figure_count$label_text("nap_intensity")` shows that while they have smaller differences than XCalibur, there is still a trend of increasing difference with the raw scan-centric intensity - NAP ratios as the number of missing scans increases.
:::

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("unassigned_upset")`.**
UpSet plot with the counts of common and specific peaks without consideration of assignments for each of scancentric, MSnbase and XCalibur generated peaks for the 1ecf (**A**) and 97lipid (**B**) samples.
:::

@@

plot_mz_diffs = function(in_peaks, limits, ylimit = NULL, inset = FALSE){
  purrr::map(c("Scan-Centric", "XCalibur", "MSnbase"), function(in_method){
    tmp_peaks = in_peaks %>%
      dplyr::filter(source2 %in% in_method)
    peak_count = purrr::map_dfr(seq(1, nrow(limits)), function(in_row){
      data.frame(x = limits[in_row, "x"],
                 y = sum(between(tmp_peaks$ppm_diff, limits[in_row, "xmin"], limits[in_row, "xmax"])))
    })
    base_plot = peak_count %>%
      ggplot(aes(x = x, y = y)) +
      geom_col() +
      labs(x = "", y = "count", subtitle = in_method)

@@

    if (!is.null(ylimit)) {
      base_plot = base_plot +
        coord_cartesian(ylim = c(NA, ylimit))
    }
    if (inset && (in_method %in% c("XCalibur", "MSnbase"))) {
      inset_plot = peak_count %>%
        ggplot(aes(x = x, y = y)) +
        geom_col() +
        labs(x = "", y = "") +
        theme(axis.text.x = element_blank(),
              axis.text.y = element_text(size = 6))
      out_plot = base_plot + inset_element(inset_plot, 0.2, 0.1, 1, 1)
    } else {
      out_plot = base_plot
    }
    out_plot
  })
}

diffs_1ecf = ass_peaks_1ecf$peaks %>%
  dplyr::mutate(source2 = dplyr::case_when(
    source %in% "scanlevel" ~ "Scan-Centric",
    source %in% "xcalibur" ~ "XCalibur",
    source %in% "msnbase" ~ "MSnbase"
  ),
    ppm_diff = abs(mz_diff / mz) * 1e6)

@@

diffs_97lipid = ass_peaks_97lipid$peaks %>%
  dplyr::mutate(source2 = dplyr::case_when(
    source %in% "scanlevel" ~ "Scan-Centric",
    source %in% "xcalibur" ~ "XCalibur",
    source %in% "msnbase" ~ "MSnbase"
  ),
    ppm_diff = abs(mz_diff / mz) * 1e6)

@@

```{r increment_p_value}
tar_load(lung_compare)
lung_compare = lung_compare %>%
  dplyr::mutate(compare_id = dplyr::case_when(
    id.test %in% "corrected" ~ "Corrected",
    id.test %in% "xcalibur" ~ "XCalibur",
    id.test %in% "msnbase" ~ "MSnbase"
  ))

@@

::: {custom-style="MDPI_3.1_text"}
For large datasets, we expect that more correct peak intensities will result in better agreement between sample normalized peak intensities within a sample class.
One way to test this is to compare p-values generated from the different intensities.
`r figure_count$label_text("p_value_comparison")` compares the -1 * Log10(p-values) from `r nrow(compare_split$Corrected)` (corrected and XCalibur) or `r nrow(compare_split$MSnbase)` (MSnbase) isotope resolved molecular formulas (IMFs) assigned by SMIRFE for the scan-centric peaks and then peaks matched and intensities extracted from the other methods.
Although all of the p-vaues are somewhat (and statistically) different from those reported by the raw intensities, there are some interesting patterns of differences.
MSnbase generated p-values show the widest distribution of differences, as well as the smallest number of peaks that are present in 50% of both the cancer and non-cancer samples.
This echoes the patterns of low overlapping assigned peaks observed in both ECF samples, where many of the scan-centric amino-acid peaks only found one match from the MSnbase peaks.
Surprisingly, the truncated log-normal distribution corrected intensities generated very different p-values compared to the raw intensities, and XCalibur showed the most agreement with the raw p-values.
We tested the statistical significance of these differences using a t-test of the actual difference in log-p-values (shown in `r figure_count$label_text("p_value_comparison")`D).
It is clear that the raw scan-centric intensities are superior.
:::

@@

long_ttest = lung_compare_ttest %>%
  dplyr::select(p.adjust, comparison, estimate, conf.low, conf.high) %>%
  tidyr::pivot_longer(cols = !comparison, names_to = "measure", values_to = "value") %>%
  dplyr::mutate(compare_id = dplyr::case_when(
    comparison %in% "corrected" ~ "Corrected",
    comparison %in% "xcalibur" ~ "XCalibur",
    comparison %in% "msnbase" ~ "MSnbase"
  ))

@@

lung_plot = (compare_plots$Corrected | compare_plots$MSnbase) / (compare_plots$XCalibur | compare_sina) +
  plot_annotation(tag_levels = "A")

@@

::: {custom-style="MDPI_3.8_bullet"}
* Scans merged and peak-list exported by XCalibur
:::

@@

::: {custom-style="MDPI_3.1_text"}
To associate the XCalibur and MSnbase peaks with the scan-centric peaks, a 4 ppm (2ppm low and high) window is calculated for the scan-centric centroid, and if any peaks are found within the window, the one with the smallest m/z difference is kept as the matching peak to the scan-centric one under consideration.
:::

>>>>

::: {custom-style="MDPI_1.7_abstract"}
**Abstract:** We present a novel, scan-centric method for characterizing peaks from direct injection multi-scan Fourier transform mass spectra of complex samples that utilizes frequency values derived directly from the spacing of raw m/z points in spectral scans.
Our peak characterization method utilizes intensity independent noise removal and normalization of scan-level data to provide a much better fit of relative intensity to natural abundance probabilities for low abundance isotopologues that are not present in all of the acquired scans.
Moreover, our method calculates both peak- and scan-specific statistics incorporated within a series of quality control steps that are designed to robustly derive peak centers, intensities and intensity ratios with their scan-level variances.
These cross-scan characterized peaks are suitable for use in our previously published peak assignment methodology, Small Molecule Isotope Resolved Formula Enumeration (SMIRFE).
:::

@@

The input peak data in these studies are derived from multi-scan direct-injection FT-MS data.
SMIRFE uses expected relationships of relative peak height or intensity to natural abundance probability (NAP) across multiple peak pairs to determine the likelihood of an assignment for the peaks in question.
In the Thermo-Fisher Orbitrap FT-MS instruments, data is acquired in microscans and scans, where each scan is an aggregate of multiple microscans.
While these micro-scan and subsequent scans are expected to be analytical replicates, spray instability and temporal delay in automatic gain control pragmatically break this expectation.
Therefore, it is advantageous to keep the data at the scan level, so that any given poor quality (i.e., "bad") scan can be removed prior to aggregating the data across scans and generating a centroided peak m/z and intensity.
If the scan level data is not processed and aggregated correctly, the final peak intensities in arbitrary units and centers in m/z will have high uncertainty (total variance) [@moseleyErrorAnalysisPropagation2013].

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("motivation")`.**
**A** Xcalibur intensities (black) and theoretical intensities based on relative NAP (red) for ECF derived threonine.
The 18O isotopologue is shifted by 0.01 m/z and the NAP peaks by 0.05 m/z to aid visualization.
**B** Theoretical NAP-based-Log10-intensities and Log10 Xcalibur intensities.
The red line indicates perfect agreement.
:::

@@

The constant point difference of frequency space is useful, because some of the subsequent steps in our workflow use sliding and tiled windows where it is assumed that the sliding windows contain the same number of data points.
The m/z point-to-point differences are not constant, but can be approximated by a LOESS linear model [@Cleveland1992loess]; however, it is exceedingly difficult to create a LOESS model with an intercept of 0.
In addition, we would also need to vary the width of sliding windows according to the m/z difference at a particular m/z based on the LOESS model.
Frequency-based points suffer none of these drawbacks, and the conversion from m/z can be derived from the raw profile level data itself, which is incredibly useful.
:::

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("mz_frequency_conversion")`.** 
**A**: Intensity vs m/z for a single peak from a single scan.
Red lines denote the difference between each adjacent pair of points, and red dots the m/z midpoint between the pair of points.
The difference divided by the midpoint is used to derive the spectral frequency values in **B**.
**B** plots the intensity vs the converted frequency points derived from **A**.
The red lines denote the difference between adjacent points, which are shown in the y-axis of **C** for this single peak.
A histogram of adjacent point differences for all points is shown in **D**.
The differences for all points in a single scan vs m/z are shown in **E**, with those differences that lie within 0.49 - 0.51 shown in red.
**F** shows the plot of derived spectral frequency vs m/z, with fitted values from the linear regression model in red.
:::

@@

```{r increment_slidingwindow}
figure_count$increment("slidingwindow_count")
figure_count$increment("percentile_variation")
tar_load(mn_ratios)
tar_load(sliding_regions)
tar_load(example_sliding_region)
tar_load(noise_plot)
```

::: {custom-style="MDPI_3.1_text"}
In a dataset of this nature, we expect that much of the data is really just noise, and does not contribute meaningfully to the analysis.
Furthermore, it is expected that noise is randomly distributed across the scans.
Therefore, if we slide a window across the data and sum the number of non-zero points in each window, we expect that most of the data we encounter is actually noise.
Subsequently, we divide the counts into tiled regions of a set size (1000 frequency in this work), and examine the 99th percentile of the sliding window counts within the tiled region.
In `r figure_count$label_text("slidingwindow_count")`A we plotted the non-zero point intensities from all scans for one tiled region 2000 points wide.
The number of non-zero points in the sliding windows across this region are shown in `r figure_count$label_text("slidingwindow_count")`B, and zoomed in in C.
These provide a measure of the non-zero point density across this region.
The blue line indicates the cumulative 99th percentile for this one region, and is high, given that it looks like there are some consistent peaks across the scans in this region.
However, examining the cumulative 99th percentiles across all the regions of the spectrum, as shown in the histogram of `r figure_count$label_text("slidingwindow_count")`F, a rather large fraction of the regions have very low cumulative 99th percentile values, with a median value of 5, which results in a cutoff value of 8 (ceiling of 5 x 1.5).
Keeping only those sliding regions with a non-zero point count greater than 8 (above the red line in `r figure_count$label_text("slidingwindow_count")`B and C) results in those sliding regions shown in D, and then the point intensities in E.
A non-zero point density based method to removing potential noise is desirable because the peak intensities across scans are not identical, as shown by the need for normalization (see **Normalization of Scans**), and the distribution of RSD values before and after normalization (see **Changes in Relative Standard Deviation**).
:::

@@

```{r slidingwindow_count, dn_id = figure_count, fig.width = 12, fig.height = 8}
(example_sliding_region | sliding_regions) + plot_annotation(tag_levels = "A")
```

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("slidingwindow_count")`.**
**A - E**: A single tiled region 2000 data points wide.
**A**: Non-zero point intensities in frequency space across all scans.
**B**: Number of non-zero points in sliding regions. Blue line indicates 99th percentile of non-zero counts in this one region.
Red line is the rounded value of median x 1.5 of 99th percentiles from **F**.
**C**: Zoomed y-axis of **B** to show the non-zero counts that are below the median cutoff.
**D**: Sliding region non-zero counts that are greater than the cutoff.
**E**: Non-zero point intensities across scans in the regions from D.
**F**: Histogram of the 99th percentile cutoffs from all of the tiled windows. Red line is the median x 1.5.
:::

::: {custom-style="MDPI_3.1_text"}
Although it appears that only a handful of potential signal regions are removed between `r figure_count$label_text("slidingwindow_count")`A and E with a cutoff of 8 non-zero points in a sliding region, the number of initial signal regions rapidly explodes as the cumulative percentile cutoff is lowered, going from thousands to tens of thousands of potential peak regions to find and fit peaks within, as shown in `r figure_count$label_text("percentile_variation")`A and B for both the amino-acid ECF and lipid samples, respectively.
:::

```{r percentile_variation, dn_id = figure_count, fig.width = 10, fig.height = 6}
noise_plot + plot_annotation(tag_levels = "A")
```
:::

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("percentile_variation")`.**
The number of initial regions as a function of the percentile cutoff used for either the AA ECF (**A**) or lipid (**B**) samples.
:::

@@

::: {custom-style="MDPI_3.1_text"}
Although many other types of mass spectrometry data suffer from a variable and noisy baseline, the scan-level profile data from the Thermo-Fisher Fusion has a baseline of 0 due to manipulations in the Thermo-Fisher firmware, making the determination of the centroided values considerably easier. 
For each region initially created, the peaks in each scan within that region can be characterized (i.e., centroiding).
For centroiding, we use a simple weighted quadratic model of log(intensity) to m/z. 
We add a small constant to enable using the zero values, and weight the values by their ratio to the most intense value, which is normally the value closest to the center of peak, helping to ensure that the **true** centroid is determined. 
From the fitted model, we can derive the centroided center and the intensity of the peak, as shown in `r figure_count$label_text("centroided_peaks")`.
:::

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("breaking_regions")`.** Splitting a single region into two regions based on the peaks that are present.
**A**: The full set of raw frequency and intensity data across all scans for the region are shown.
Each horizontal trace are the point intensities in a single scan.
Clearly the region has two separate peaks within it.
**B**: The peak centroids (frequency and intensity) for each peak in black.
The tiled regions (red) are used to quantify the number of peaks.
**C**: The number of peaks within each tiled region shown as a histogram.
Each group of non-zero adjacent regions will be merged to form a new peak region.
:::

@@

::: {custom-style="MDPI_3.1_text"}
Due to differences in how many ions are captured in the trap and the limited dynamic range of the detector, the observed peak intensities for the same analyte vary between scans (see `r figure_count$label_text("breaking_regions")`A and B for example).
Using the median peak differences between scans, it is possible to normalize the peak intensities across scans.
However, there are two issues with these peak intensities across scans: (a) some peak intensities correlate with the scan number (i.e., scan acquisition order); and (b) some peak differences between scans are correlated with intensity.
The solution to (a) is to do a two-pass normalization.
After the first pass, the peaks whose intensity is correlated with scan order are detected (absolute Pearson correlation > 0.5).
In the second pass, the correlated peaks are removed, and normalization is carried out again.
`r figure_count$label_text("intensity_scan")`A shows an example peak whose intensity across scans is correlated with scan number.
The solution to (b) (peak differences correlated with intensity) is to only use the most intense peaks, as shown in `r figure_count$label_text("intensity_scan")`B.
The highlighted peaks in `r figure_count$label_text("intensity_scan")`B are those with an intensity greater than 0.7 of the maximum intensity observed in that scan, and at least visually, their differences are **not** correlated with intensity.
If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in `r figure_count$label_text("normalization_factors")`A and B.
Though the two-pass normalization (doublenorm) and single-pass normalization using the most intense peaks (singlenorm_int) generate very similar normalization factors (`r figure_count$label_text("normalization_factors")`B and C), which is expected as they both use the most intense peaks. 
But removing the scan-correlated peaks does change the majority of the normalization factors.
:::

@@

::: {custom-style="MDPI_3.1_text"}
We have previously described the presence of high peak density (HPD) artifacts in FT-MS spectra [@mitchellNewMethodsIdentify2018].
Ideally, the peak characterization procedure should reduce and mitigate their presence and effect in the resultant reported peaks.
Their presence should be minimized by removing noise peaks, and removing peak regions that have multiple reported peaks from the same scan.
However, we expect they may also present as characterized peaks that have larger than expected frequency level standard deviations (FSD) when calculated across scans.
These peaks can be detected by simply examining the distribution of FSDs and marking outlier peaks.
To verify the mitigation of HPD regions, we converted centroided m/z's from Xcalibur to frequency values using the previously calculated model for that sample, and measured peak density to detect HPD regions, and compare them with the scan-centric peaks and FSD outliers.
Xcalibur peak density was measured using a sliding window ten points wide and a stride of one point.

@@

::: {custom-style="MDPI_3.1_text"}
Each step in the peak characterization either changes the overall number of peaks coming from each scan (sliding windows and breaking initial regions) or the overall intensity of the points within a scan.
Therefore, one way to quantify any potential *improvements* in the characterized peaks is to look at the relative standard deviation (RSD) for the characterized scan level peak intensities (calculated as the standard deviation of peak heights across scans divided by the mean peak height) and compare them as each processing step is introduced.
`r figure_count$label_text("rsd_method")` illustrates the peak height RSD distributions for four different samples.
Up to two modes are reported for each distribution, where a mode location is only reported if it is ≥ 0.2X the most intense mode.
For samples 1ecf and 2ecf, there is a general shift in the RSD to the left, going from the bottom processing methods (msnbase_only) to the top processing method (filtersd), representing a visually clear improvement.
For samples 49lipid and 97lipid, the improvements are visually more subtle, having a slight shift to the left as well as a narrowing and smoothing of the RSD distribution.
Part of this subtlety is likely due to the bi-modality of these two distributions.
This is not surprising, since these two samples are non-polar extractions from tissue and are biochemically more complex.
`r table_count$label_text("rsd_method")` provides more quantitative metrics.
At two decimal places, the filtersd, doublenorm, and singlenorm processing methods give superior and nearly identical results for three of the four samples, especially in terms of mean and median RSD.
However, for sample 1ecf, the max RSD is much higher for singlenorm, highlighting its instability.
:::

@@

::: {custom-style="MDPI_3.1_text"}
The 97lipid filtersd RSD distribution (as well as others) is bi-modal.
`r figure_count$label_text("rsd_intensity")` plots each peaks RSD as a function of the mean intensity across scans, for every peak present in at least three scans (top), or the peaks present in at least 80% of the scans (bottom).
At least part of the bi-modality in the RSD distribution appears to be related to this dependence of RSD on intensity, as well as truncation as peaks appear in fewer scans.
This is supported by part of the distribution disappearing when we require that peaks be in at least 80% of the available scans (difference between `r figure_count$label_text("rsd_intensity")` top and bottom).

@@

::: {custom-style="MDPI_3.1_text"}
Although we expected that the corrected scan-centric intensities would behave similarly or better compared to the raw intensities, `r supp_figure_count$label_text("nap_intensity")` shows that while they have smaller differences than Xcalibur, there is still a trend of increasing difference with the raw scan-centric intensity - NAP ratios as the number of missing scans increases.
:::

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("unassigned_upset")`.**
UpSet plot with the counts of common and specific peaks without consideration of assignments for each of scancentric, MSnbase and Xcalibur generated peaks for the 1ecf (**A**) and 97lipid (**B**) samples.
The black points with connected vertical lines identify which set intersections are represented in the bar at the top.
A single black point identifies what was specific to the set and not in any other sets.
:::

@@

plot_mz_diffs = function(in_peaks, limits, ylimit = NULL, inset = FALSE){
  purrr::map(c("Scan-Centric", "Xcalibur", "MSnbase"), function(in_method){
    tmp_peaks = in_peaks %>%
      dplyr::filter(source2 %in% in_method)
    peak_count = purrr::map_dfr(seq(1, nrow(limits)), function(in_row){
      data.frame(x = limits[in_row, "x"],
                 y = sum(between(tmp_peaks$ppm_diff, limits[in_row, "xmin"], limits[in_row, "xmax"])))
    })
    base_plot = peak_count %>%
      ggplot(aes(x = x, y = y)) +
      geom_col() +
      labs(x = "", y = "count", subtitle = in_method)

@@

    if (!is.null(ylimit)) {
      base_plot = base_plot +
        coord_cartesian(ylim = c(NA, ylimit))
    }
    if (inset && (in_method %in% c("Xcalibur", "MSnbase"))) {
      inset_plot = peak_count %>%
        ggplot(aes(x = x, y = y)) +
        geom_col() +
        labs(x = "", y = "") +
        theme(axis.text.x = element_blank(),
              axis.text.y = element_text(size = 6))
      out_plot = base_plot + inset_element(inset_plot, 0.2, 0.1, 1, 1)
    } else {
      out_plot = base_plot
    }
    out_plot
  })
}

diffs_1ecf = ass_peaks_1ecf$peaks %>%
  dplyr::mutate(source2 = dplyr::case_when(
    source %in% "scanlevel" ~ "Scan-Centric",
    source %in% "xcalibur" ~ "Xcalibur",
    source %in% "msnbase" ~ "MSnbase"
  ),
    ppm_diff = abs(mz_diff / mz) * 1e6)

@@

diffs_97lipid = ass_peaks_97lipid$peaks %>%
  dplyr::mutate(source2 = dplyr::case_when(
    source %in% "scanlevel" ~ "Scan-Centric",
    source %in% "xcalibur" ~ "Xcalibur",
    source %in% "msnbase" ~ "MSnbase"
  ),
    ppm_diff = abs(mz_diff / mz) * 1e6)

@@

```{r increment_p_value}
tar_load(lung_compare)
lung_compare = lung_compare %>%
  dplyr::mutate(compare_id = dplyr::case_when(
    id.test %in% "corrected" ~ "Corrected",
    id.test %in% "xcalibur" ~ "Xcalibur",
    id.test %in% "msnbase" ~ "MSnbase"
  ))

@@

::: {custom-style="MDPI_3.1_text"}
For large datasets, we expect that more correct peak intensities will result in better agreement between sample normalized peak intensities within a sample class.
One way to test this is to compare p-values generated from the different intensities.
`r figure_count$label_text("p_value_comparison")` compares the -1 * Log10(p-values) from `r nrow(compare_split$Corrected)` (corrected and Xcalibur) or `r nrow(compare_split$MSnbase)` (MSnbase) isotope resolved molecular formulas (IMFs) assigned by SMIRFE for the scan-centric peaks and then peaks matched and intensities extracted from the other methods.
Although all of the p-vaues are somewhat (and statistically) different from those reported by the raw intensities, there are some interesting patterns of differences.
MSnbase generated p-values show the widest distribution of differences, as well as the smallest number of peaks that are present in 50% of both the cancer and non-cancer samples.
This echoes the patterns of low overlapping assigned peaks observed in both ECF samples, where many of the scan-centric amino-acid peaks only found one match from the MSnbase peaks.
Surprisingly, the truncated log-normal distribution corrected intensities generated very different p-values compared to the raw intensities, and Xcalibur showed the most agreement with the raw p-values.
We tested the statistical significance of these differences using a t-test of the actual difference in log-p-values (shown in `r figure_count$label_text("p_value_comparison")`D).
It is clear that the raw scan-centric intensities are superior.
:::

@@

long_ttest = lung_compare_ttest %>%
  dplyr::select(p.adjust, comparison, estimate, conf.low, conf.high) %>%
  tidyr::pivot_longer(cols = !comparison, names_to = "measure", values_to = "value") %>%
  dplyr::mutate(compare_id = dplyr::case_when(
    comparison %in% "corrected" ~ "Corrected",
    comparison %in% "xcalibur" ~ "Xcalibur",
    comparison %in% "msnbase" ~ "MSnbase"
  ))

@@

lung_plot = (compare_plots$Corrected | compare_plots$MSnbase) / (compare_plots$Xcalibur | compare_sina) +
  plot_annotation(tag_levels = "A")

@@

::: {custom-style="MDPI_3.8_bullet"}
* Scans merged and peak-list exported by Xcalibur
:::

@@

::: {custom-style="MDPI_3.1_text"}
To associate the Xcalibur and MSnbase peaks with the scan-centric peaks, a 4 ppm (2ppm low and high) window is calculated for the scan-centric centroid, and if any peaks are found within the window, the one with the smallest m/z difference is kept as the matching peak to the scan-centric one under consideration.
:::
